---
title: "Interim Project"
author: "Tim Long, Todd Dearing, Charles"
date: "October 17, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Import test and train data sets.
test <- read.csv("C:/Predictive Analytics Assignments/FinalProject/test.csv")
train <- read.csv("C:/Predictive Analytics Assignments/FinalProject/train.csv")

library("knitr", lib.loc="~/R/win-library/3.3")
library("arm", lib.loc="~/R/win-library/3.3")
library("caret", lib.loc="~/R/win-library/3.3")

areaModel <- standardize(lm(SalePrice~LotArea, data=train))
display(areaModel)
plot(train$LotArea, train$SalePrice, main=paste ("Plot of", names(train$LotArea)))
plot(areaModel)
hist(train$LotArea)
```
# Introduction

This project will use the Ames housing data set of 1460 different house prices. The data will be used to predict house prices for the test set of 1459 houses using 5 described variables. The project will present Linear Models demonstrating the selection criteria for the five specific variables and how they perform. A description of the process of cleaning data, variable selection, and how the different models performed on the train and test set is provided. At the end of this project, the RMSE of a prediction of the sales price for the test set of the data using the best model will be given. 

# Data and Cleaning

Inspection of the data and data descriptions found that minimal data cleaning was needed to produce a good prediction model. When looking at a plot of LotArea with Sales as the outcome variable it was concluded that Lot size had 5 outliers that skewed the data and decreased the predictive power that LotArea had to offer. After removal of the outliers and implementing that into a new data frame, LotArea was an effective predictive factor in models that utilized it. 

```{r, include=FALSE}
nLotArea <- train[which(train$LotArea <= 70000),]
set.seed(512)
Indexes <- sample(nrow(nLotArea), 0.7 * nrow(nLotArea), replace = F)
tr <- nLotArea[Indexes, ]
tt <- nLotArea[-Indexes, ]
```
# Model and Model Development

Given the context of the problem and the data available it was determined that the model developed to describe the housing sale prices in Iowa should be simple, use relevant variables, and accurate.  This model will be used to describe what variables affect home sale prices in Iowa to the layman.  A limit was imposed upon the model that it shouldn’t include more than five variables.  The five variable limit will reinforce the simplicity of the model as well as meet the requirements set in the project description.  

Relevant variables were determined to be those that would be easily understood and commonly used by an individual with basic understanding of real estate and home ownership. The variables selected to create the model should require little to no explanation.  This will support the simplicity and usability of the model.

Finally the model should be accurate and resistant to errors.  A linear model was accepted to fit these requirements.  Linear models are simple and therefore easy to explain, visualize, and interpret.  Linear models are common, resilient, need little background knowledge to apply, and require less computational power compared to other modeling methods.  The model name and the RMSE are reported below for each model.

An initial linear model using “Neighborhood” as the only variable was created.  This was used as a baseline to compare further models.  The categorical variable Neighborhood is easy to understand and it contains a geographical component that groups homes according to location.  Home prices aren’t independent of each other and the Neighborhood variable is a simple way to include this behavior and provide a degree of normalization.  
```{r, include=FALSE}
#Model with only Neighboorhood (subset according to lotarea <= 70000)

display(hoodPrice <- lm(SalePrice~Neighborhood, data=nLotArea))

nLotArea$SalePred <- predict(hoodPrice, nLotArea)
```
Neighborhood only model:
```{r}
RMSE(nLotArea$SalePred, nLotArea$SalePrice)
```
Above ground living area (“GrLivArea”) was then added as a variable.  Above ground living area is often used in home sales adds to describe a home and has an impact on the value of a home.  
```{r, include=FALSE}
#Model with only Neighborhood and GrLivArea (subset according to lotarea <= 70000)

display(hoodGr <- lm(SalePrice~Neighborhood+GrLivArea, data=nLotArea))

nLotArea$SaleGrPred <- predict(hoodGr, nLotArea)
```
Neighborhood and above ground living area model:
```{r}
RMSE(nLotArea$SaleGrPred, nLotArea$SalePrice)
```
Lot area (“LotArea”) was added to the linear model because it was a variable typically used when describing a property.  It is straightforward and easy to explain.
```{r, include=FALSE}
#Model with Neighborhood, GrLivArea, and LotArea (subset according to lotarea <= 70000)

display(hoodGrLa <- lm(SalePrice~Neighborhood+GrLivArea+LotArea, data=nLotArea))

nLotArea$SaleGrLaPred <- predict(hoodGrLa, nLotArea)
```
Neighborhood, above ground living area, and lot area model:
```{r}
RMSE(nLotArea$SaleGrLaPred, nLotArea$SalePrice)
```
The age of the home was calculated (2011 - YearBuilt) and added to the model because it significantly increased the fit of the linear model to the data. 
```{r, include=FALSE}
#Model with Neighborhood, GrLivArea, age, and LotArea (subset according to lotarea <= 70000)

nLotArea$age <- (2011-nLotArea$YearBuilt)

display(hoodGrLaYr <- lm(SalePrice~Neighborhood+GrLivArea+LotArea+age, data=nLotArea))

nLotArea$SaleGrLaYrPred <- predict(hoodGrLaYr, nLotArea)
```
Neighborhood, above ground living area, age, and lot area model:
```{r}
RMSE(nLotArea$SaleGrLaYrPred, nLotArea$SalePrice)
```
Finally overall quality of the the home (“OverallQual”) was added to the linear model.  The condition of the home carries a great deal of influence when deciding to purchase a home and how much to purchase it for.  It was added since it improved the fit of the model and wouldn’t require additional explanation to the typical homebuyer.  
```{r, include=FALSE}
#Model with Neighborhood, GrLivArea, YearBuilt, OverallQual, and LotArea (subset according to lotarea <= 70000)

display(hoodGrLaYrQ <- lm(SalePrice~Neighborhood+GrLivArea+LotArea+age+OverallQual, data=nLotArea))

nLotArea$SaleGrLaYrQPred <- predict(hoodGrLaYrQ, nLotArea)
```
Neighborhood, above ground living area, age, overall quality, and lot area model:
```{r}
RMSE(nLotArea$SaleGrLaYrQPred, nLotArea$SalePrice)
```
```{r, include=FALSE}
#Run predictions on the test set.  

#Report RMSE (maybe need to write our own function for RMSE since the standard R function returns 'NA'.

#Report standard error of residuals

#Create table presenting results of all models used in development for comparison. (Create dataframe)
Model <- c("hoodPrice", "hoodGr ", "hoodGrLa", "hoodGrLaYr", "hoodGrLaYrQ" )
RMSE <- c(RMSE(nLotArea$SalePred, nLotArea$SalePrice),RMSE(nLotArea$SaleGrPred, nLotArea$SalePrice) , RMSE(nLotArea$SaleGrLaPred, nLotArea$SalePrice),RMSE(nLotArea$SaleGrLaYrPred, nLotArea$SalePrice), RMSE(nLotArea$SaleGrLaYrQPred, nLotArea$SalePrice)) 
R_sq <- c(summary(hoodPrice)$r.squared, summary(hoodGr)$r.squared, summary(hoodGrLa)$r.squared, summary(hoodGrLaYr)$r.squared, summary(hoodGrLaYrQ)$r.squared) 
Adj_R_sq <- c(summary(hoodPrice)$adj.r.squared, summary(hoodGr)$adj.r.squared, summary(hoodGrLa)$adj.r.squared, summary(hoodGrLaYr)$adj.r.squared, summary(hoodGrLaYrQ)$adj.r.squared) 
```
```{r, include=FALSE}
test$age <- (2011 - test$YearBuilt)

tt$age <- (2011 - tt$YearBuilt)

tt$salesPred <- predict(hoodGrLaYrQ, newdata = tt)
```
# Results

The table below contains the models that were used in the development process as well as indicators of their accuracy. 

```{r, include=FALSE}
df = data.frame(Model, RMSE, R_sq, Adj_R_sq)
```
```{r}
kable(df, digits = 3, caption = "Model Results")
```

The accuracy of the model increased with the addition of each variable.  The final model had the lowest RMSE and residual standard deviation for the models applied to the training data set.  The final model fit the data well and the included variables can describe a majority of the data.  
```{r}
plot(hoodGrLaYrQ$residuals, main = "Residual Plot")
```
Finally a the train data was separated into train and test sets to validate assumptions made about the model accuracy.  Below is the RMSE of the test set when the model generated by the train data was used on the test set. The RMSE compares well to the RMSE originally reported when the model was developed.
```{r}
RMSE(tt$SalePrice, tt$salesPred)
```
There are improvements that can be made to this model.  Although there was consistency in the predictions, the standard deviation of the residuals was still significant enough to reduce the predictive accuracy of the model.  The selected variables LotArea, Age, Overall Quality, Above Ground Living Area, and Neighborhood are easy to understand, but interactions weren't considered and should be investigated.  
```{r}
plot(hoodGrLaYrQ$fitted.values, hoodGrLaYrQ$residuals, main = "Residuals vs Fitted")
```
There is some nonlinearity in the data that can be seen when the residuals are plotted against the fitted values.  The addition of a higher degree polynomial variable improve the overall fit of the model.

Ultimately the suggested linear model is a acceptable descriptor of the Iowa home sales price.  The model is simple, employed relevant variables, and maintained a consistent accuracy when applied to the test data.  Interactions and higher degree polynomials should be considered when trying to improve the fit of the model in the future. 
